Explicit B-spline Regularization in Diffeomorphic Image Registration

### 2013-10-03 paper.tex
ENH:  Adding Log jacobians.

\begin{abstract}
Important methodological developments in the evolution of image registration algorithms include those 
in which the correspondence relationship is characterized by diffeomorphisms.
The  popularity of these approaches is due largely to their 
topological properties and success in providing biologically plausible 
solutions to small and large deformation 
estimation problems. Variants of diffeomorphic algorithms include those characterized by 
time-varying and constant velocity fields, and symmetrical considerations. 
Prior information in the form of regularization is used to enforce transform plausibility taking the form of physics-based constraints or through some approximation thereof, e.g. Gaussian smoothing (a la Thirion's Demons \citep{thirion1998}).  In the context of the original Demons' framework, the traditional free-form deformation method can be viewed as a variant in which explicit regularization is achieved through the B-spline basis functions.  
This characterization can be used to provide alternative ``flavors'' of diffeomorphic image registration solutions with several advantages which we describe in this work.  Implementation is open source and available through the Insight Toolkit and our Advanced Normalization Tools (ANTs) repository.  A thorough evaluation with the well-known SyN algorithm \citep{avants2008}, implemented within the same framework, and its B-spline analog is performed using open data and open source evaluation tools.
\tiny
\keyFont{ \section{Keywords:} 
open science best practices comparison reproducibility software }
\end{abstract}

@@

Establishment of anatomical and functional correspondence
is a crucial step towards insight into certain biological 
sciences.  Neuroscience research efforts, such as characterizing 
brain morphology, require accurate and robust methods for
producing such mappings.  The 
extensive literature detailing methodology is evidence of the rich history of 
algorithmic development which continues contemporaneously.
We highlight several key contributions which are particularly relevant to the work presented.

Free-form deformation (FFD) image registration, preceded by related work for
geometric modeling \citep{sederberg1986}, originated with such important
contributions as \cite{szeliski1997,thevenaz1998,rueckert1999}.  Continued 
development within this early spline-based paradigm produced additional innovations
such as integrated similarity metrics \citep[e.g.][]{mattes2003}, additional transformation
constraints \citep[e.g.][]{rohlfing2003}, and open source contributions \citep[e.g.][]{ibanez2005,klein2010,shackleford2010}.

Parallel to this branch of algorithmic progress are the informally 
denoted ``dense transforms''
perhaps best exemplified by Thirion's seminal contribution \citep{thirion1998}.
Relationships with earlier elastic \citep{bajcsy1989,gee1993} and fluid \citep{christensen1996} registration methods are elucidated in 
the works of \cite{bro-nielsen1996,pennec1999} who observe that
Gaussian smoothing, characteristic of Demons, of the update or displacement
field is a greedy approximation for solving the partial differential equations governing
the physics of an elastic or fluid deformation.  However, the use of such 
approximations comes with the caveat that physical properties, such as topological
regularity,
are no longer guaranteed.

It is interesting to note that within this context, traditional FFD algorithms
can be viewed as a type of fluid-like Demons approach 
where, rather than produced by Gaussian convolution, smooth fields are produced using B-splines for roughly approximating the solution to the viscous fluid system.  This was hinted at in our earlier work \citep{tustison2009} where we showed that fitting the update field to a B-spline object using a fast approximation routine \citep{tustison2006} is equivalent to a preconditioning of the standard gradient used in gradient descent-based FFD 
optimization.  Further details can be gleaned from the ITK implementation of this work%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'') 
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

Continuing from the work of \cite{christensen1996} and subsequent exploration into the mathematical formalisms of diffeomorphisms \citep[e.g.][]{dupuis1998}, 
the well-known Large Deformation Diffeomorphic Metric Mapping (LDDMM) algorithm  
was proposed in \cite{Beg2005}.  In contrast to the mapping produced
by \cite{christensen1996}, LDDMM yields the geodesic solution in the space of diffeomorphisms between two images.

>>>>

\begin{abstract}
Important methodological developments in the evolution of image registration algorithms include those 
in which the correspondence relationship is characterized by diffeomorphisms.
The  popularity of these approaches is due largely to their 
topological properties and success in providing biologically plausible 
solutions to small and large deformation 
estimation problems. Popular variants of diffeomorphic algorithms include those characterized by 
time-varying and constant velocity fields, and symmetrical considerations. 
Prior information (i.e. regularization) is used to enforce transform plausibility taking the form of physics-based constraints or through some approximation thereof, e.g. Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In the context of the original Demons' framework, the traditional free-form deformation method can be viewed as a variant in which explicit regularization is achieved through the B-spline basis functions.  
This characterization can be used to provide alternative B-spline ``flavored'' diffeomorphic image registration solutions with several advantages which we describe in this work.  Implementation is open source and available through the Insight Toolkit and our Advanced Normalization Tools (ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm \citep{avants2008}, implemented within the same framework, and its B-spline analog is performed using open data and open source evaluation tools.
\tiny
\keyFont{ \section{Keywords:} 
open science, best practices, reproducibility, software }
\end{abstract}

@@

Establishment of anatomical and functional correspondence
is a crucial step towards gaining insight into certain biological 
sciences.  Neuroscience research efforts, such as characterizing 
brain morphology, require accurate and robust methods for
producing such mappings.  The 
extensive literature detailing methodology is evidence of the rich history of 
algorithmic development which continues contemporaneously.
We highlight several key contributions which are particularly relevant to the work presented.

Free-form deformation (FFD) image registration, characterized by  regularization based on the B-spline basis functions, has several 
advantages including ease of algorithmic understanding, 
good performance, desirable geometric properties, and several available 
implementations.  Current research was preceded by related work for
geometric modeling \citep{sederberg1986} and originated with such important
contributions as \cite{szeliski1997,thevenaz1998,rueckert1999}.  
Continued development within this early spline-based paradigm produced additional innovations such as integrated similarity metrics \citep[e.g.][]{mattes2003}, additional transformation constraints \citep[e.g.][]{rohlfing2003}, and notable open source implementations \citep[e.g.][]{ibanez2005,klein2010,shackleford2010}.

Parallel to this branch of algorithmic progress are the informally 
denoted ``dense transforms''
perhaps best exemplified by Thirion's seminal contribution \citep{thirion1998}.
Relationships with earlier elastic \citep{bajcsy1989,gee1993} and fluid \citep{christensen1996} registration methods are detailed in 
the works of \cite{bro-nielsen1996} and \cite{pennec1999} who observe that
Gaussian smoothing, characteristic of Demons, of the update or displacement
field is a greedy approximation for solving the partial differential equations governing
the physics of an elastic or fluid deformation, respectively.  However, the use of such 
approximations entails that physical properties, such as topological
regularity, are no longer guaranteed.

It is interesting to note that within this context, traditional FFD algorithms
can be viewed as a type of fluid-like Demons approach 
where, rather than produced by Gaussian convolution, regularized fields are produced using B-splines for roughly approximating the solution to the viscous fluid system.  This was hinted at in our earlier work \citep{tustison2009} where we showed that fitting the update field to a B-spline object using a fast approximation routine \citep{tustison2006} is equivalent to a preconditioning of the standard gradient used in gradient descent-based FFD optimization.  As mentioned in \cite{tustison2009}, this preconditioning is used to reduce the hemstitching induced by the traditional gradient-based FFD formulation.
Further details can be gleaned from the ITK implementation of this 
earlier work%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'') 
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

Continuing from the work of \cite{christensen1996} and subsequent exploration into the mathematical formalisms of diffeomorphisms \citep[e.g.][]{dupuis1998}, 
the well-known Large Deformation Diffeomorphic Metric Mapping (LDDMM) algorithm  
was proposed in \cite{Beg2005}.  In contrast to the mapping produced
by \cite{christensen1996}, LDDMM yields the geodesic solution in the space of diffeomorphisms between two images. Since its introduction, LDDMM has
inspired much innovation in the image registration literature.  
Symmetrical considerations in the velocity field parameterization, for example, are discussed in \cite{avants2008} in the context of a cross correlation similarity metric which minimizes the bias of the resulting transformation in selecting the identity of the fixed and moving images.  A greedy version of this algorithm which has proven successful in general application \citep{Klein2009}.
